\toolnameend: A Tool for Visualizing Patterns of Spreadsheet Function Combinations

### 2016-04-26 doc/paper.tex
Got paper to 10 pages.

	Alongside the tool, we discuss several of the key decisions we faced during
	design, such as what to do with redundant formulae in a sheet or how to
	represent optional arguments, and we describe how our overall design goals both
	helped and hindered this process. We then apply the tool to a number of case
	studies to evaluate how much it could contribute to several different contexts,
	most notably code smell detection and spreadsheet education.

	To better understand how people might use \toolnameend, we also conducted a
	user study to judge how usable and intuitive the tool actually is. This grounds
	the discussion of its potential uses while also underscoring some of its
	inherent limitations and setting the way for future extensions and improvements
	upon the concept.

@@

	\begin{figure}[h] \centering \includegraphics[scale=.5]{start} \caption{How the
			IF tree looks at the start} \label{fig:startpic} \end{figure}

@@

	Knowing that it is the first argument which contains the conditionals, we click
	on the square labeled ``1" to explore. To save space, when more than 10 unique
	arguments have been observed in any position, the tool displays only the first
	ten, with an option to display the rest by clicking the arrow
	\includegraphics{glossary-arrow}.

@@

	We see that IF contains a number of comparison operators as its first argument,
	such as = and $\le$, and boolean-returning functions, like ISNA and AND, with
	simple equality being the most common and some use as ABS being the least seen
	of everything actually used. \par

	From here, we can further explore the common options among these functions.
	Clicking on the ``=" node will yield two arguments and expanding each of them
	will peer into the range of common values of equality comparison, which Figure
	\ref{fig:fullpic} shows. \par

@@

	From these images, we can provide a quick glossary of glyphs within the
	visualization to aid interpretation. \begin{itemize} \item
		\includegraphics{glossary-green} - \textbf{Function nodes} are green circles
		labeled with the function name. Their size is determined by their relative
		frequency, scaled logarithmically, to the frequency of the root node in the
		tree. They may have any number of arguments, including zero, as determined by
		the API.

		\item \vspace{.25cm}\includegraphics{glossary-leaf} - \textbf{Value nodes} are
		white circles with a generic label. They represent elements of a formula which
		accept no arguments, like references (A1), ranges (A1:A10), numbers, boolean
		values, strings, and errors. However, \toolname treats still functions with
		zero arguments, like NOW, as function nodes.

@@

		\item \textit{Presentation}: The JSON files, meanwhile, feed into the
		presentation code, implemented in Javascript with much help from the
		visualization library D3\footnote{https://d3js.org/}. We chose D3 primarily
		for its accessibility: given the JSON, the visualization can be simply
		embedded into a webpage accessible through a browser. \end{itemize} A
	demonstration of the tool can be found at
	https://github.com/DeveloperLiberationFront/Excel-Function-Visualizer.

@@

	Guided by the goals in Section \ref{goals}, we faced a number of decisions in
	how to represent the data in the visualization. Below, I've outlined those that
	had the largest impact in the values that appear in the tree. \begin{itemize}

@@

	\begin{figure}[h] \centering \includegraphics[width=.5\textwidth]{lookup-2}
		\includegraphics[width=.5\textwidth]{lookup-3} \caption{LOOKUP with two and
			three arguments}  \label{fig:lookups} \end{figure}

@@

	\begin{figure}[h] \centering \includegraphics[width=.5\textwidth]{vlookup}
		\includegraphics[width=.5\textwidth]{hlookup} \caption{VLOOKUP and HLOOKUP with
			four arguments} \label{fig:vhlookups}\end{figure}

@@

	Additionally, \toolname offers a simple way of finding examples where certain
	functions were used together. Though a standard text-search tool might yield
	the same results for a function alone, combinations can require more complex
	text queries, whereas here, the information is already captured in nodes.  \par

	However, we found that even though a user can explore all of a function's
	configurations, this does not mean that they will learn exactly what it does.
	P1, examining unfamiliar functions like KURT and PV, could describe precisely
	what types of arguments it accepted and how many. But even after exploring the
	spreadsheets for context, they could not accurately describe what the functions
	produced without consulting the official documentation. Whether through
	reticence of the tool or obscurity of the functions themselves, the exploratory
	environment is not enough to teach users \textit{when} to use a function.
	However, we could address this in future designs, perhaps, by linking directly
	to such documentation from within the tool without violating any design
	concerns in Section \ref{goals}. \par

	\subsection{Study Limitations} Though we did not start the study by telling
	that the data source came from Enron, two users studied the file names to
	deduce its roots in finance, if not Enron itself. Because of these notorious
	connotations, then, they might have been inclined from this information to
	orient their exploration around finance-related functions, potentially limiting
	their findings or perceived users for the tool. Though, one also said that,
	without this essential context, they could not have properly explored a dataset
	in the first place. Comments like these point out the trade-offs of unguided
	exploration, too: though it might not restrict them to a single purpose, it
	might, at the same time, not afford them any mindset at all with which to
	interpret the data. \par

	Furthermore, the data which users explored was processed and created before we
	had fully completed the tool, meaning some of the counts and patterns in the
	trees were inaccurate. However, we decided that this was not a pressing
	concern, being that we were primarily concerned with whether any interesting
	reports could be made at all, not with whether these reports were 100\%
	accurate assessments of reality. \par

	Throughout the study, we noticed that users misunderstood some of the icons
	and, as such, their early discoveries were inaccurate. For example, P1
	interpreted the numbers on the argument boxes to mean that all of the box's
	children had exactly that many arguments -- that is, clicking on the dark blue
	``1" would return all the single-argument functions instead of the first
	argument of a function -- and P2 initially took the solid lines to mean
	different paths in formula construction rather than different arguments of the
	same function. Though the users naturally did not explicitly report
	misinterpreting the icons, these misunderstandings became clear in their early
	evaluations of the spreadsheet practice. Though we clarified the meanings when
	these misunderstandings became apparent, these events point to a failure in the
	tool to properly indicate every function with clarity.   \par

	\section{Discussion} We began these studies with the intention of evaluating
	our progress towards the goals in Section \ref{goals}. Working from these
	results, we noted many places where the tool aligned well or fell short.

@@

	Participants also asked for more options to quickly navigate the tree. For
	example, P2 asked for the ability to expand the entire tree at once and icons
	that better communicated the presence of optional arguments and their
	quantities. The more familiar participants grew with the tool, the faster they
	wanted to explore, yet it supported expert navigation up to the point of
	expanding all of a node's children up to one level. \par

	Other than that, we did avoid designing anything that intentionally evaluated
	formula designs. Nevertheless, \toolname inadvertently emphasizes bad design of
	certain types, such as all of those in Section \ref{badsmells} marked by their
	visual signatures. Even so, our participants found the discovery of bad smell
	to be the most interesting aspect about it, to the point where they could
	imagine business scenarios in which micromanagers use it to ensure precise
	design parameters across a department's spreadsheets! To design in that
	direction, however, would mean a reevaluation of this design goal, since
	intentionally guiding users toward known smells may guide them away from new,
	unmarked ones.

@@

	We realized early that some of the popular functions would present a problem:
	some would have thousands of variations in large datasets, and showing all of
	these variations at once would hopelessly overwhelm the user. We therefore made
	a number of choices, such as the expansion arrows and the use of generic types
	in place of specific primitive values, that would reduce the variation to
	something more manageable.

	Nevertheless, during the user study, P3 remarked that they specifically avoided
	the most popular functions, anticipating a flood of nodes from which they could
	salvage nothing interesting -- which might not have been a bad prediction,
	considering that the SUM is built from 910 function and argument nodes stemming
	the root and IF, at least 18000. P2 corroborated this by explaining how they
	gravitated toward exploring specific examples in particular value nodes over
	navigating the full crowd of the tree, but they complicate it further by also
	saying that the tool at its best when it shows either every possibility of an
	argument side-by-side (for comparison) or nothing on that branch at all, a
	balance that is hard to strike with node-dense trees.

@@

	\subsection{Informative Qualification} As we found in the case study, the tool
	excels at making formula examples available throughout a tree. Though the tool
	cannot outright explain what a formula does function-by-function, users
	nevertheless found it valuable for the way it offers them with numerous
	practical examples of functions previously unknown or confusing to them,
	especially within the LOOKUP family. In the user study (Section
	\ref{sec:userstudy}), furthermore, we also observed how a participant used the
	repeated instance of a spreadsheet author's name to identify a potential
	problem with a particular designer rather than a design itself, demonstrating
	its use on several levels. \par

	However, the benefits described in the previous sections imply a hidden
	complication: if the tool educates through example but does not explicitly
	distinguish between good design and bad design, how can we be sure that people
	will not inadvertently learn bad design from it? In the user study, we
	discussed its use in education and smell detection by bringing up the issues
	surrounding the LOOKUP functions; this discussion, though, presupposes a
	distinguishing eye that knew which examples practiced good design and which
	exhibited the common errors. But a beginner might not know the difference.
	Robillard and Deline, in their study, remarked that good API examples should
	demonstrate best practices, but, right now, we have no way to assure this with
	\toolname. As with the problems in achieving unbiased exploration, correcting
	for this danger may entail a reevaluation of design goals.

@@

	\begin{figure} \centering \includegraphics[width=.5\textwidth]{index}
		\caption{A difficult question: how often is MATCH used in both the second and
			third arguments?} \label{fig:index} \end{figure}

>>>>

	Alongside the tool, we discuss several of the key decisions we faced during
	design, such as what to do with redundant formulae in a sheet or how to
	represent optional arguments, and we describe how our overall design goals both
	helped and hindered this process. We then apply the tool to a number of case
	studies to evaluate how much it could contribute to several different contexts,
	most notably code smell detection and spreadsheet education. To better understand how people might use \toolnameend, we also conducted a
	user study to judge how usable and intuitive the tool actually is. This grounds
	the discussion of its potential uses while also underscoring some of its
	inherent limitations and setting the way for future extensions and improvements
	upon the concept.

@@

	\begin{figure}[h] \centering \includegraphics[scale=.4]{start} \caption{How the
			IF tree looks at the start} \label{fig:startpic} \end{figure}

@@

	Knowing that it is the first argument which contains the conditionals, we click
	on the square labeled ``1" to explore. To save space, when more than 10 unique
	arguments were found in a position, the tool displays only the first
	ten, with an option to display the rest by clicking the arrow
	\includegraphics{glossary-arrow}.

@@

	We see that IF contains a number of comparison operators as its first argument,
	such as = and $\le$, and boolean-returning functions, like ISNA and AND, with
	simple equality being the most common and some use as ABS being the least seen
	of everything actually used. From here, we can further explore the common options among these functions.
	Clicking on the ``=" node will yield two arguments and expanding each of them
	will peer into the range of common values of equality comparison, which Figure
	\ref{fig:fullpic} shows. \par

@@

	From these images, we can provide a quick glossary of glyphs within the
	visualization to aid interpretation. \begin{itemize} \item
		\includegraphics{glossary-green} - \textbf{Function nodes} are green circles
		labeled with the function name. Their size is determined by their relative
		frequency, scaled logarithmically, to the frequency of the root node in the
		tree. They may have any number of arguments, including zero, as determined by
		the API. Users can select any of these nodes to view the examples from the 
		provided dataset that use this function in the node's position in the tree.

		\item \vspace{.25cm}\includegraphics{glossary-leaf} - \textbf{Value nodes} are
		white circles with a generic label. They represent elements of a formula which
		accept no arguments, like references (A1), ranges (A1:A10), numbers, boolean
		values, strings, and errors. However, \toolname treats still functions with
		zero arguments, like NOW, as function nodes. Like function nodes, these also 
		contain examples.

@@

		\item \textit{Presentation}: The JSON files, meanwhile, feed into the
		presentation code, implemented in Javascript with much help from the
		visualization library D3\footnote{https://d3js.org/}. We chose D3 primarily
		for its accessibility: given the JSON, the visualization can be simply
		embedded into a webpage accessible through a browser. \end{itemize} The code and a
	demonstration of \toolname can be found at
	https://github.com/DeveloperLiberationFront/Excel-Function-Visualizer.

@@

	Guided by the goals in Section \ref{goals}, we faced a number of decisions in
	how to visualize the data. Below, we've outlined those that
	had the largest impact in the values that appear in the tree. \begin{itemize}

@@

	\begin{figure}[h] \centering \includegraphics[scale=.75]{lookup-2}
		\includegraphics[scale=.75]{lookup-3} \caption{LOOKUP with two and
			three arguments}  \label{fig:lookups} \end{figure}

@@

	\begin{figure}[h] \centering \includegraphics[scale=.75]{vlookup}
		\includegraphics[scale=.75]{hlookup} \caption{VLOOKUP and HLOOKUP with
			four arguments} \label{fig:vhlookups}\end{figure}

@@

	Additionally, \toolname offers a simple way of finding examples where certain
	functions were used together, demonstrated when P4 took special interest in how people
	frequently used MATCH as the second argument of INDEX or how they used date functions
	like DATE and MONTH as lookup values in VLOOKUP. Though a standard text-search tool might yield
	the same results for a function alone, combinations can require more complex
	text queries, whereas here, the information is already captured in nodes.  \par

	However, we found that even though a user can explore all of a function's
	configurations, this does not mean that they will learn exactly what it does.
	P1, examining unfamiliar functions like KURT and PV, could describe precisely
	what types of arguments it accepted and how many. But even after exploring the
	spreadsheets for context, they could not accurately describe what the functions
	produced without consulting the official documentation. Whether through
	reticence of the tool or obscurity of the functions themselves, the exploratory
	environment is not enough to teach users \textit{when} to use a function.
	However, we could address this in future designs, perhaps, by linking directly
	to such documentation from within the tool without violating any design
	concerns in Section~\ref{goals}. \par

	\subsection{Study Limitations} Though we did not start the study by telling
	that the data source came from Enron, two users studied the file names to
	deduce its roots in finance, if not Enron itself. Because of these notorious
	connotations, then, they might have been inclined from this information to
	orient their exploration around finance-related functions, potentially limiting
	their findings or perceived users for the tool. However, one participant also said that,
	without this essential context, they could not have intelligently explored a dataset
	in the first place. Comments like these point out the trade-offs of unguided
	exploration, too: though it might not restrict them to a single purpose, it
	might, at the same time, not afford them any mindset at all with which to
	interpret the data. \par

	Furthermore, the data which users explored was processed and created before we
	had fully ensured that the tool could extract all of the useful data from the 
	spreadsheets, meaning some of the counts and patterns in the
	trees were inaccurate. However, we decided that this was not a pressing
	concern, being that we were primarily concerned with whether any interesting
	reports could be made at all, not with whether these reports were 100\%
	accurate assessments of reality. \par

	Throughout the study, we noticed that users misunderstood some of the icons
	and, as such, their early discoveries were inaccurate. For example, P1
	interpreted the numbers on the argument boxes to mean that all of the box's
	children had exactly that many arguments -- that is, clicking on the dark blue
	``1" would return all the single-argument functions instead of the first
	argument of a function -- and P2 initially took the solid lines to mean
	different paths in formula construction rather than different arguments of the
	same function. Though the users naturally did not explicitly report
	misinterpreting the icons, these misunderstandings became clear in their early
	evaluations of the spreadsheet practice. Though we clarified the meanings when
	these misunderstandings became apparent, these events point to a failure in the
	tool to indicate every function clearly.   \par

	\section{Limitations} We began these studies with the intention of evaluating
	our progress towards the goals in Section \ref{goals}. Working from these
	results, we noted many places where the tool aligned well or fell short.

@@

	Other than that, we did avoid designing anything that intentionally evaluated
	formula designs. Nevertheless, \toolname inadvertently emphasizes bad design of
	certain types, such as all of those in Section~\ref{badsmells} marked by their
	visual signatures. Even so, our participants found the discovery of bad smell
	to be the most interesting aspect about it, to the point where they could
	imagine business scenarios in which micromanagers use it to ensure precise
	design parameters across a department's spreadsheets! To design in that
	direction, however, would mean a reevaluation of this design goal, since
	intentionally guiding users toward known smells may guide them away from new,
	unmarked ones.

@@

	We realized early that some of the popular functions would have thousands of variations in large datasets, and showing all of
	these variations at once would hopelessly overwhelm the user. We therefore made
	a number of choices, such as the expansion arrows and the use of generic types
	in place of specific primitive values, that would reduce the variation to
	something more manageable.

	Nevertheless, during the user study, P3 remarked that they specifically avoided
	the most popular functions, anticipating a flood of nodes from which they could
	salvage nothing interesting. P2 corroborated this by explaining how they
	gravitated toward exploring specific examples in particular value nodes over
	navigating the full, crowded tree, but they complicate it further by also
	saying that the tool at its best when it shows either every possibility of an
	argument side-by-side (for comparison) or nothing on that branch at all, a
	balance that is hard to strike with node-dense trees.

@@

	\subsection{Informative Qualification} As we found in the case study, the tool
	excels at making formula examples available throughout a tree. 	
	However, the benefits described in the previous sections imply a hidden
	complication: if the tool educates through example but does not explicitly
	distinguish between good design and bad design, how can we be sure that people
	will not inadvertently learn bad design from it? In the user study, we
	discussed its use in education and smell detection by bringing up the issues
	surrounding the LOOKUP functions; this discussion, though, presupposes a
	distinguishing eye that knew which examples practiced good design and which
	exhibited the common errors. A beginner might not know the difference.
	Robillard and Deline, in their study, remarked that good API examples should
	demonstrate best practices, but, right now, we have no way to assure this with
	\toolname. As with the problems in achieving unbiased exploration, correcting
	for this danger may entail a reevaluation of design goals.

@@

	\begin{figure} \centering \includegraphics[scale=.75]{index}
		\caption{A difficult question: how often is MATCH used in both the second and
			third arguments?} \label{fig:index} \end{figure}
